---
title: "Epimodel"
author: "Reed Scott"
date: "2024-11-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(vegan)
source('Occu_abun_practice.R')
```

# Introduction

I am assuming that you have already read the document on simulating abundance of species in a metacommunity. As
a reminder, our goal is to understand whether or not landscape level R~0~ is impacted by $\beta$ diversity. We
have simulated a dataset with abundances for each species. Now we need to see how the abundance of each species
and the connectivty of patches interact to impact prevalence of the disease.

# Initializing

The first step is to determine initial conditions. We need to determine the amount of susceptible individuals 
and infected individuals at the start of this simulation. For now, I have assumed that the most abundant 
species (spp1) is the most susceptible (highest probability of getting disease) while the least abundant spp 
(spp6) is the most resistant. Because this uses an SIS model, the number of infectious individuals I can be 
determined by the total population - S. Total population is also just S+I. Finally we have to determine a
starting disease level. In this case, we are using the amphibian chytrid fungus as an example, so I have
chose what I *think* is a reasonable starting number of 1000 zoospores.

```{r, initial conditions}
S <- ceiling(meta_comm1[,1:6]*c(0.8,0.85,0.9,0.93,0.95,.99)) #starting value of susceptibles
I <- meta_comm1[,1:6] - S #starting value of infecteds
N <- S+I
Z <- 1000 #starting amount of zoospores
```

# Species characteristics

Ater choosing the initial values, we must assign species specific traits to determine how their abundance will impact disease risk. These include a birth rate (b), death rate (d), tranmission coefficient(beta), recovery rate (v), dispersal rate (phi), and BD load / shedding rate (lambda). These are shown in the table below.

```{r species characteristics, echo = FALSE}
Species <- c("Spp1","Spp2","Spp3","Spp4","spp5","spp6")
b <- c(0.6,0.5,0.4,0.3,0.2,0.1) #host birth rate
d <- c(0.06,0.05,0.04,0.03,0.02,0.01) #host death rate
beta <- c(0.00013,0.00012,0.00011,0.00010,0.00009,0.00008) #transmission per species
v <- c(0.4,0.5,0.6,0.7,0.8,0.9) #recovery rate
phi <- c(0.09,0.08,0.07,0.06,0.05,0.04) #dispersal rate
lambda <- c(3.0,2.5,2.0,1.5,1.0,0.5) #log(bd) load. losely based off figure 2 from wilber 2020 
species_chara <- data.frame(Species = Species,
                            birth = b, 
                            death = d, 
                            trans = beta, 
                            recovery = v, 
                            dispersal = phi, 
                            shedding = lambda)
species_chara
```

*Note: right now, these characteristics are relatively arbitrary. I came up with values that I think would be*
*in the range of plausible values. However, these are essentially made up. A key next step is to find more*
*realistics values. This may include asking Mark what values they observed in their study.*

Lastly, we must set some more characteristics about the meta-community within which we are operating.
We must establish gamma, or the decay rate of zoospores of Bd. We have to establish a matrix, c, of the 
connectivty values between patches. And we must establish the area of each patch. We also have to establish
the length of time over which this will occur. In this case, I have set time = 90, to represent a potential
90 day field season.

```{r, meta-comm chara}
# meta-community characteristics
gamma <- 0.001 #zoospore decay rate
c <- matrix(data = rnorm(n = num_patches^2, mean = 0.5, sd = 0.1),
            nrow = num_patches, 
            ncol = num_patches)
#Connectivity of patches
A <- rnorm(n = num_patches, mean = 0.5, sd = 0.1) #area ratios
time <- 90 #how many "days" do I want in the season
```

*Note: right now, c is a matrix of random values with mean 0.5 and sd = 0.1. This will likely need to be*
*adjusted considerably in my opinion. We may need to set distance values between each pair of patches, and*
*then consider how distance affects connectivty. Additionally, we have to consider that within this matrix,*
*along the diagonal is the probability that an individual stays in the same patch. There will likely be a*
*different formulation for this probability.*

# Running the simulation over time

Alright, we're finally at the main event! This is a relatively short bit of code, but it does a lot.

First, we'll want to establish an empty list, of length time. We will use this list to record the number
susceptible individuals of each species, infectious individuals of each species, total number of individuals
of each species, and the time (i.e. "day"). 

With that set we can start the for loop. The for loop runs for time t. There are 3 key steps to this loop.

1. Calculate delta_s, delta_I, and delta_Z. This is the number of new susceptible and ifnectious individuals
   for each species, as well as the number of zoospores being added to the pool.
2. Calculate the new S, I, Z, and N values. This just simply adds the delta values to their respective
   initial values to calculate the "new" values for S,I,Z, and N.
3. Package it all in a list. We will do this so that we get a nice output we can later use for analysis.

```{r simulation}
pop_list <- vector("list", length = time)
for (t in 1:time) {
  delta_s <- b*N - d*S - beta*S*Z + v*I + phi*sum(-c*S + c*S) #for now have excluded area of patches
  #may want to add that back in though
  delta_I <- beta*S*Z-(v+d)*I+phi*sum(-c*I + c*I)#for now have excluded area of patches
  #may want to add that back in though
  delta_Z <- sum(lambda*I - gamma*Z)
  S <- S+delta_s
  I <- I+delta_I
  z <- Z+delta_Z
  N <- S + I
  t = t
  pop <- list(Susceptible = S, Infectious = I, Zoospores = Z, Total = N, Time = t)
  pop_list[[t]] <- pop
}
```

I am not going to display pop_list because it's huge: it's 90 lists, each of which contain 5 dataframes. 
Additionally, right now the output is unrealistic. However, I think that as we fine tune the parameters,
we will find this to be a good functional model.

# Conclusion

I believe what I have now is solid progress and a really good start to the simulation! However, I have no
delusions. There is still a lot of work to be done one this. Most important, I believe, is that I need to
tune the parameters to be more biologicall relevant / realistic. I also still need to consider how I will
be analyzing the output of this SIS model. Additionally, the goal is to simulate multiple different
metacommunities and compare landscape R~0~ values, which I have not yet done. So, yes, there is still a lot
of work to do. I am nonetheless excited at the progress I've made.