---
title: "Meta-disease update"
author: "Reed Scott"
institute: "University of Vermont"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["custom.css"]
    includes:
    after_body: insert-logo.html
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/gitrepos/MetaDisease/ReportsAndUpdates/01.15.2024")
setwd("D:/gitrepos/MetaDisease/ReportsAndUpdates/01.15.2024")
library(vegan)
library(tidyverse)
library(ggplot2)
library(RColorBrewer)
library(patchwork)
library(gt)
set.seed(1234)
```

# Goals for today?

1. Background
2. Establishing meta-communities
3. Running simulations

---

# Background
## General Refresher

* Previous studies have shown that biodiversity is an important part disease dynamics
--

* However, none have looked at these within the context of a meta-community
--

* We assert that the *difference* between communities ($\beta diversity$) may play an important role in disease dynamics
--

* In this project we will be building on prior work done by Mark, which established methods for calculating a landscape level $R_0$
---

# Background
## Study system

* Originally, I had suggest this use a "fake system" with species that acted as charicatures of real species
--

* However, I have found that difficult
--

* Instead, I will focus on a simulation of a real system. I will use 6 real species and 1 real disease. Those are:
  - *P. regilla*
  - *A. boreas*
  - *T. Granulosa*
  - *T. torosa*
  - *R. catesbianus*
  - *R. draytonii*
  - And the amphibian chytrid fungus
* To do this, I will have to use real world parameters of these species, which I will get into later.
---

# Background
## Hypotheses

* Our goal: to understand the how a diverse array of communities can impact disease risk
--

* Our specific hypothesis: as $\beta$ diversity increases, landscape level $R_0$ should decrease

---

# Establishing meta-communities

* Up until now, I feel I've largely glossed over how I formed the simulated metacommunities I want to use for this study.
* I think it's about time I described them in a decent amount of detail, so here goes.

---

# Establishing meta-communities

First, we set up the number of possible species, number of patches, $\psi$ for each species, and rank abundance

```{r Init meta params}
set.seed(1234)
num_patches <- 5 #number of patches in metacommunity
num_spp <- 6 #number of POSSIBLE spp in metacommunity
S <- c(0.83, #PREG
       0.62, #TGRAN
       0.14, #TTOR
       NA, #ABOR
       0.12, #RCAT
       0.64) #RDRAY
S[4] <- runif(n = 1, min = S[5], max = S[3])
K <- c(10,6,5,4,3,2) 
```

---

# Establishing meta-communities
Then we simulate 100 different meta-communities.
I will explain in depth how I did that

```{r build metas}
N <- 100 #number of metacommunity simulations to run
meta_comm_list <- vector("list",N)
beta_list <- vector("list", N)
nestedness_list <- vector("list", N)
for(n in 1:N){
  meta_comm <- data.frame(matrix(NA, nrow = num_patches, ncol = num_spp))
  for(i in 1:nrow(meta_comm)){
    for(j in 1:ncol(meta_comm)){
      #determine occurrence and abundance of spp 1
      meta_comm[i,1] <- rbinom(n = 1, size = 1, prob = S[1])
      meta_comm[i,1] <- ifelse(meta_comm[i,1] == 1, K[1],0)
      
      #determine occurrence and abundance of spp2
      meta_comm[i,2] <- rbinom(n = 1, size = 1, prob = S[2])
      meta_comm[i,2] <- ifelse(meta_comm[i,2] == 1, K[2],0)
      
      #determine occurrence and abundance of spp 3
      meta_comm[i,3] <- rbinom(n = 1, size = 1, prob = S[3])
      meta_comm[i,3] <- ifelse(meta_comm[i,3] == 1, K[3],0)
      
      #determine occurrence and abundance of spp 4
      meta_comm[i,4] <- rbinom(n = 1, size = 1, prob = S[4])
      meta_comm[i,4] <- ifelse(meta_comm[i,4] == 1, K[4],0)
      
      #determine occurrence and abundance of spp 5
      meta_comm[i,5] <- rbinom(n = 1, size = 1, prob = S[5])
      meta_comm[i,5] <- ifelse(meta_comm[i,5] == 1, K[5],0)
      
      #determine occurrence and abundance of spp 6
      meta_comm[i,6] <- rbinom(n = 1, size = 1, prob = S[6])
      meta_comm[i,6] <- ifelse(meta_comm[i,6] == 1, K[6],0)
    }
  }
  meta_comm$Patch <- c('Patch1','Patch2','Patch3','Patch4','Patch5')
  colnames(meta_comm) <- c('Spp1','Spp2','Spp3','spp4','spp5','Spp6','PatchID')
  beta_list[[n]] <- mean(betadiver(meta_comm[,1:6], method = 'w'))
  nestedness_list[[n]] <- nestedtemp(comm = meta_comm[,1:6])[7]
  meta_comm_list[[n]] <- meta_comm
}
```

---

class: center, middle

# And voila! Just like that we have 100 simulated meta-communities.
# Now, let's get in to the epidiomoligcal modeling!

---

# The epi. model
## The necessary functions

* This are taken from the .py code provided by Mark. I just translated it to R.
* There are 3 key functions
1 "build_R_with_data" builds the necessary R matrix
--

```{r build_R_with_data}
build_R_with_data <- function(R0s,b,beta,S,P) {
  fullR <- matrix(data = 0, nrow = P*S, ncol = P*S)
  for (p in 1:P) {
    tR0 <- R0s[p,]
    Rmat <- matrix(data = rep(unlist(tR0),S), nrow = S, ncol = S)
    tb <- b[p,]
    bmat <- matrix(data = rep(unlist(tb),S), nrow = S, ncol = S)
    # I have removed the following code (in .py) but should ask mark if I need to add something for
    # Force of infection
    # tλ = λs[:, p]
    # λmat = np.repeat(tλ, S).reshape(S, S).T
    # λ_ratios = λmat / λmat.T    
    Rmat <- Rmat * bmat
    Rmat[is.nan(Rmat)] = 0
    Rmat[is.infinite(Rmat)] = 0
    Rmat <- as.matrix(Rmat)
    start <- ifelse(p == 1, p,(S*p)-(S-1))
    stop <- start + S - 1
    fullR[start:stop, start:stop] <- Rmat
  }
  
  return(fullR)
}
```

---
# The epi. model
## The necessary functions

* This are taken from the .py code provided by Mark. I just translated it to R.
* There are 3 key functions
2 "build_B_with_data" builds the necessary B matrix

```{r build_B_with_data}
build_B_with_data <- function(Cmat, phi, b, S, P) {
  Cmat_axis <- colSums(Cmat)
  diag_list <- vector("list",P^2)
  diag_list_names <- array(dim = c(sqrt(length(diag_list)), sqrt(length(diag_list))))
  for (p in 1:P) { #loop over columns
    for (j in 1:P) { #loop over rows
      diag_list_names[j,p] <- paste(p,j,sep="_")
    }
  }
  diag_list_names <- as.vector(diag_list_names)
  names(diag_list) <- diag_list_names
  for (p in 1:P) { #loop over columns
    for (j in 1:P) { #loop over rows
      tZ <- matrix(0,nrow = S, ncol = S)
      new_diag <- array(dim = S)
      new_diag <- if(p == j){
        (-1)*b[p,] - phi*Cmat_axis[p]
        } else{
          phi*Cmat[j,p]
          }
      diag(tZ) <- new_diag
      x <- paste(p,j,sep="_")
      diag_list[[x]] <- tZ 
    }
  }
  diag_matrix <- matrix(diag_list, nrow = P, ncol = P)
  tB <- list()
  for (j in 1:P) {
    tB[[j]] <- matrix(data = unlist(diag_matrix[j,]), nrow = S*P, ncol = S, byrow = T) #looks like this worked!
  }
  
  B <- do.call(cbind, tB[1:P]) #need to determine if this is what B should look like
  return(B) 
}
```

---

# The epi. model
## The necessary functions

* This are taken from the .py code provided by Mark. I just translated it to R.
* There are 3 key functions
3 "landscape_R0" The max eigen value of K is landscape $R_0$

```{r landscape_R0}
landscape_R0 <- function(R0_spps,Cmat,b, beta, phi,S,P){
  R <- build_R_with_data(R0s = R0_spps,
                         b = b,
                         beta = beta,
                         S = S,
                         P = P)
  B <- build_B_with_data(Cmat = Cmat,
                         phi = phi,
                         b = b,
                         S = S,
                         P = P)
  K <- R * solve(-1*B)
  return(list(K = K, R = R, B = B))
}

```

---

# The epi. model
## Establishing species parameters

* We need to establish some relevant information about each species in order to build our models.
--

* Specifically we need
  - Birth rate
  - Death rate
  - Recovery rate
  - Dispersal rate
  - transmission rate
  - Connectivity between patches
  
---

# The Epi model
## Birth rate

* This is pretty simple
* Since we are looking within a season, I set birth rate = 0 for all species

---
# The Epi model
## Death rate

Species        | death rate (d)
---------------|---------------------------------------------------
P. regilla     | 0.024
A. boreas      | 0.0029
L. catesbeianus| 0.144
L. draytonii   | 0.04
T. granulosa   | Between P. regilla and A. boreas
T. torosa      | Between P. regilla and A. boreas
---

# The Epi model
## Recovery rate

* Finding reliable information on the recovery rate of adults has proven... difficult
* I found 1 study and for the rest I assigned random values that I believe are ecologically relevant

Species        | recovery rate (b)
---------------|---------------------------------------------------
P. regilla     | Should be lowest
A. boreas      | Should be lower than L. catesbeianus
L. catesbeianus| 0.012
L. draytonii   | Should be highest
T. granulosa   | Because it is a newt, should be relatively high
T. torosa      | Because it is a newt, should be relatively high

---

# The Epi model
## Transmission rate

* $\beta$ is... hard. and worth talking about for a minute
* Few studies I have found directly state a value for $\beta$
* So I have elected to use ecological context and "fill in the blanks"
* For this part I make 2 key assumptions
  1. The most abundant species will have the highest $\beta$
  2. $\beta_{ss}$ will likely be higher than $\beta_{si}$
  
* I will spare you the code the create the necessary S*S $\beta$ matrix, but I do want to show the resulting matrix and graph

---
# The Epi model
## Transmission rate (cont'd)

```{r hidden params, echo = FALSE}
Species <- c("PREG","TGRAN","TTOR","ABOR","RCAT","RDRAY")
# PREG = Pseudacris Regilla (Pacific tree frog)
# ABOR = Anaxyrus boreas (western toad)
# RCAT = Rana catesbeiana (American bullfrog)
# RDRAY = Rana draytonii (Califronia red legged frog)
# TTOR = Taricha torosa (California newt)
# TGRAN = Taricha granulosa (rough-skinned newt)

birth <- rep(0, times = 6) #currently have birth rate set to 0 b/c this is a within season model
# if we change to a multi-season model, will need to add in some birth rate
# b <- c(0.6,0.5,0.4,0.3,0.2,0.1)/time #host birth rate
d <- c(0.024,
       NA,
       NA,
       0.0029,
       0.0144,
       0.04)
d[2] <- runif(n = 1, min = d[[4]], max = d[[1]])       
d[3] <- runif(n = 1, min = d[[4]], max = d[[2]])       
# Citations for death rate
# PREG = Jameson 1956
# ABOR = Pilliod 2010
# RCAT = Howell 2020
# RDRAY = Feller 2017
# TTOR = johnson 2013 + assumptions of dilution effects
# TGRAN =  johnson 2013 + assumptions of dilution effects
v <- c(NA,
       NA,
       NA,
       NA,
       0.012460223,
       NA)
# Citation for RCAT recovery = Daszak 2004
low_recovery <- runif(n = 4, min = 0, max = v[5])
v[1] <- min(low_recovery)
v[4] <- max(low_recovery)
mid_low_recovery <- low_recovery[low_recovery != v[1] & low_recovery != v[4]]
v[2] <- min(mid_low_recovery)
v[3] <- max(mid_low_recovery)
v[6] <- runif(n = 1, min = v[5], max = 0.1) #should find citation for max recovery rate
v

#Determining dispersal rate
psi <- c(0.83, 0.62, 0.14,NA, 0.12,0.64)
psi[4] <- runif(n = 1, min = psi[5], max = psi[3])
```

```{r trans rate, echo=FALSE, include = FALSE, results ='hide'}
x <- seq(0,1,length = 100)
trans_rate <- function(n = 1,x = seq(0,1,length = 100),a,b){
  trans <- rbeta(n = n, shape1 = a, shape2 = b)
  plot(x = seq(0,1, length.out = 100), y = dbeta(x, shape1 = a, shape2 = b))
  return(trans)
}

# PREG
intra_PREG <- trans_rate(a = 4,b = 2)
inter_PREG <- trans_rate(a = 4, b = 2.5)

plot(x = x,dbeta(x = x,shape1 = 4,shape2 = 2), ylab = "density", type = 'l', col = "red")
lines(x=x, dbeta(x, shape1 = 4, shape2 = 2.5), col = "blue")


# TGRAN
intra_TGRAN <- trans_rate(a = 4, b = 2.25)
inter_TGRAN <- trans_rate(a = 4, b = 2.5)

plot(x = x,dbeta(x = x,shape1 = 4,shape2 = 2.25), ylab = "density", type = 'l', col = "red")
lines(x=x, dbeta(x, shape1 = 4, shape2 = 2.5), col = "blue")

#TTOR
intra_TTOR <- trans_rate(a = 3, b = 2.5)
inter_TTOR <- trans_rate(a = 3, b = 2.75)

plot(x = x,dbeta(x = x,shape1 = 3,shape2 = 2.5), ylab = "density", type = 'l', col = "red")
lines(x=x, dbeta(x, shape1 = 3, shape2 = 2.75), col = "blue")

#ABOR
intra_ABOR <- trans_rate(a = 2.5, b = 2.75)
inter_ABOR <- trans_rate(a = 2.5, b = 3.0)

plot(x = x,dbeta(x = x,shape1 = 2.5,shape2 = 2.75), ylab = "density", type = 'l', col = "red")
lines(x=x, dbeta(x, shape1 = 2.5, shape2 = 3.0), col = "blue")

#RCAT
intra_RCAT <- trans_rate(a = 2.0, b = 3.0)
inter_RCAT <- trans_rate(a = 2.0, b = 3.25)

plot(x = x,dbeta(x = x,shape1 = 2.0,shape2 = 3.0), ylab = "density", type = 'l', col = "red")
lines(x=x, dbeta(x, shape1 = 2.0, shape2 = 3.25), col = "blue")

#RDRAY
intra_RDRAY <- trans_rate(a = 1.5, b = 3.25)
inter_RDRAY <- trans_rate(a = 1.5, b = 3.5)

plot(x = x,dbeta(x = x,shape1 = 1.5,shape2 = 3.25), ylab = "density", type = 'l', col = "red")
lines(x=x, dbeta(x, shape1 = 1.5, shape2 = 3.5), col = "blue")
```

```{r beta matrix, echo=FALSE}
beta <- matrix(data = NA, nrow = num_spp, ncol = num_spp)
for (i in 1:nrow(beta)) {
  for (j in 1:ncol(beta)) {
    beta[i,j] <- if(i == 1 & j == 1){
      intra_PREG}else if(i != 1 & j == 1){
        inter_PREG} else if(i == 2 & j == 2){
          intra_TGRAN} else if(i != 2 & j == 2){
            inter_TGRAN} else if(i == 3 & j == 3){
              intra_TTOR} else if(i != 3 & j == 3){
                inter_TTOR} else if(i == 4 & j == 4){
                  intra_ABOR} else if(i != 4 & j == 4){
                    inter_ABOR} else if(i == 5 & j == 5){
                      intra_RCAT} else if(i != 5 & j ==5){
                        inter_RCAT} else if(i == 6 & j == 6){
                          intra_RDRAY} else if(i != 6 & j == 6){
                            inter_RDRAY}
  }
}
beta <- beta/90
gt(data.frame(beta))
```
---

# Connectivty of Patches

* I assumed  that all patches had the same connectivity
* I also assumed that the likelihood of staying was higher than the likelihood of moving between patches.
* This resulted in the following matrix:
--

```{r connectivity, echo = FALSE}
#Connectivity of patches
stay <- rbeta(n = 1, shape1 = 4, shape2 = 2) #probability individuals stay in a patch?
go <- rbeta(n = 1, shape1 = 1.5, shape2 = 3.5) # probability individuals move
c <- matrix(data = NA,
            nrow = num_patches, 
            ncol = num_patches)
for(i in 1:ncol(c)){
  for(j in 1:nrow(c)){
    c[i,j] <- ifelse(i == j, stay, go)
  }
}
gt(data.frame(c))
```

---

# The Simulation

* Okay, so everything is coming to a head.
--

* We have built the meta-communities, and established the necessary parameters.
--

* Now it's time to read it all into a simulation
--

```{r The simul}

# Simulation over metacommunities
result2 <- data.frame(matrix(data = NA, nrow = length(meta_comm_list), ncol = 18))
colnames(result2) <- c("TotalAbundance","PREG","TGRAN","TTOR","ABOR","RCAT","RDRAY",
                       "PREG_rel","TGRAN_rel","TTOR_rel","ABOR_rel","RCAT_rel","RDRAY_rel",
                       "BetaDiversity","Gamma_diversity",
                       "Beta_relative","LandscapeR0", "MetaCommID")
for (a in 1:length(meta_comm_list)) {
  S <- ceiling(meta_comm_list[[a]][,1:6]*c(0.8,0.85,0.9,0.93,0.95,.99)) #value of susceptibles
  I <- meta_comm_list[[a]][,1:6] - S #value of infecteds
  N <- S+I #total pop of a patch
  S <- as.matrix(S)
  I <- as.matrix(I)
  N <- as.matrix(N)
  N_meta <- colSums(N)
  phi <- ifelse(N_meta > 0, psi/N_meta, 0) #get dispersal rate
  r0_species_patch <- matrix(nrow = num_patches, ncol = num_spp) #create empty matrix, later will calc R0
  b <- matrix(nrow = num_patches, ncol = num_spp) #create empty matrix, later will calc b
  for (p in 1:num_patches) {
    for(q in 1:num_patches){
      for (i in 1:num_spp) {
        for (j in 1:num_spp) {
          #establish parameters for time t of species s
          connectivity_s <- phi[i]*sum(-c[p,q]*S[p,i] + c[q,p]*S[q,i])
          connectivity_I <- phi[i]*sum(-c[p,q]*I[p,i] + c[q,p]*I[q,i])
          birth_rate <- birth[i]*N[p,i]
          death_s <- d[i]*S[p,i]
          loss_I <- (v[i]+d[i])*I[p,i]
          FI <- ifelse(N[p,j] > 0, beta[i,j]*S[p,i]*(I[p,j]/N[p,j]), 0)
          
          #change in rate of susceptible + infectious individuals
          
          #calculate R0 for each species at each patch
          b[p,i] <- v[i] + d[i] #gets loss rate
          r0_species_patch[p,i] <- beta[i,i]/b[p,i]
        }
      }
    }
  }
  
  result2[a,1] <- sum(N) #total abundance
  #abdunance of each species
  for (i in 2:(1+num_spp)) {
    result2[a,i] <- sum(N[,i-1])
    result2[a,(i+num_spp)] <- result2[a,i]/result2[a,1]
  }
  
  #add beta diversity
  beta_diversity <- betadiver(N, method = 'w')
  result2[a,14] <- mean(beta_diversity, na.rm =T) #beta diversity
  result2[a,15] <- sum(ifelse(result2[a,1:6] > 0, 1,0)) #gamma diversity
  
  result2[a,16] <- result2[a,7]/result2[a,8] #beta diversity / total abundance

    #calculate landscape R0
  r0_landscape <- landscape_R0(R0_spps = r0_species_patch,
                               Cmat = c,
                               b = b,
                               beta = beta,
                               phi = phi,
                               S = num_spp,
                               P = num_patches)
  
  result2[a,17] <- eigen(r0_landscape[[1]])$values[1] #landscape R0
  result2[a,18] <- a #metacommunity ID
}
```

---

# The Simulation
## Results

```{r sim output, echo = FALSE}
gt(head(result2))
```
        
---

class: center, middle
# Are you excited? Because I'm excited
## Let's look at some graph

---

# Graphing

.pull-left[
* First, I wanted to check the relationship between $\beta$ diversity and $R_{0L}$
]

```{r beta graph, echo = F, message=FALSE}
beta_plot <- ggplot(data = result2, aes(x = BetaDiversity, y = LandscapeR0))+
  geom_point()+
  geom_smooth(method = 'lm')+
  theme_classic()+
  labs(title = 'Effect of Beta diversity on Landscape R0')
beta_plot
```
---

# Graphing
* Hmm, not what we predicted. I wonder if abundance has an affect

```{r abund graph, echo = F, message=FALSE}
abun_R0_plot <- ggplot(data = result2, aes(x = TotalAbundance, y = LandscapeR0))+
  geom_point()+
  geom_smooth(method = 'lm')+
  theme_classic()+
  labs(title = 'Effect of Abundance on Landscape R0')
abun_R0_plot
```
