---
title: "Meta-disease update"
author: "Reed Scott"
institute: "University of Vermont"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["custom.css"]
    includes:
    after_body: insert-logo.html
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "D:/gitrepos/MetaDisease/ReportsAndUpdates/01.15.2024")
setwd("D:/gitrepos/MetaDisease/ReportsAndUpdates/01.15.2024")
library(vegan)
```

# Goals for today?

1. Background
2. Establishing meta-communities
3. Running simulations

---

# Background
## General Refresher

* Previous studies have shown that biodiversity is an important part disease dynamics
--

* However, none have looked at these within the context of a meta-community
--

* We assert that the *difference* between communities ($\beta diversity$) may play an important role in disease dynamics
--

* In this project we will be building on prior work done by Mark, which established methods for calculating a landscape level $R_0$
---

# Background
## Study system

* Originally, I had suggest this use a "fake system" with species that acted as charicatures of real species
--

* However, I have found that difficult
--

* Instead, I will focus on a simulation of a real system. I will use 6 real species and 1 real disease. Those are:
  - *P. regilla*
  - *A. boreas*
  - *T. Granulosa*
  - *T. torosa*
  - *R. catesbianus*
  - *R. draytonii*
  - And the amphibian chytrid fungus
* To do this, I will have to use real world parameters of these species, which I will get into later.
---

# Background
## Hypotheses

* Our goal: to understand the how a diverse array of communities can impact disease risk
--

* Our specific hypothesis: as $\beta$ diversity increases, landscape level $R_0$ should decrease

---

# Establishing meta-communities

* Up until now, I feel I've largely glossed over how I formed the simulated metacommunities I want to use for this study.
* I think it's about time I described them in a decent amount of detail, so here goes.

---

# Establishing meta-communities

First, we set up the number of possible species, number of patches, $\psi$ for each species, and rank abundance

```{r Init meta params}
set.seed(1234)
num_patches <- 5 #number of patches in metacommunity
num_spp <- 6 #number of POSSIBLE spp in metacommunity
S <- c(0.83, #PREG
       0.62, #TGRAN
       0.14, #TTOR
       NA, #ABOR
       0.12, #RCAT
       0.64) #RDRAY
S[4] <- runif(n = 1, min = S[5], max = S[3])
K <- c(10,6,5,4,3,2) 
```

---

# Establishing meta-communities
Then we simulate 100 different meta-communities.
I will explain in depth how I did that

```{r build metas}
N <- 100 #number of metacommunity simulations to run
meta_comm_list <- vector("list",N)
beta_list <- vector("list", N)
nestedness_list <- vector("list", N)
for(n in 1:N){
  meta_comm <- data.frame(matrix(NA, nrow = num_patches, ncol = num_spp))
  for(i in 1:nrow(meta_comm)){
    for(j in 1:ncol(meta_comm)){
      #determine occurrence and abundance of spp 1
      meta_comm[i,1] <- rbinom(n = 1, size = 1, prob = S[1])
      meta_comm[i,1] <- ifelse(meta_comm[i,1] == 1, K[1],0)
      
      #determine occurrence and abundance of spp2
      meta_comm[i,2] <- rbinom(n = 1, size = 1, prob = S[2])
      meta_comm[i,2] <- ifelse(meta_comm[i,2] == 1, K[2],0)
      
      #determine occurrence and abundance of spp 3
      meta_comm[i,3] <- rbinom(n = 1, size = 1, prob = S[3])
      meta_comm[i,3] <- ifelse(meta_comm[i,3] == 1, K[3],0)
      
      #determine occurrence and abundance of spp 4
      meta_comm[i,4] <- rbinom(n = 1, size = 1, prob = S[4])
      meta_comm[i,4] <- ifelse(meta_comm[i,4] == 1, K[4],0)
      
      #determine occurrence and abundance of spp 5
      meta_comm[i,5] <- rbinom(n = 1, size = 1, prob = S[5])
      meta_comm[i,5] <- ifelse(meta_comm[i,5] == 1, K[5],0)
      
      #determine occurrence and abundance of spp 6
      meta_comm[i,6] <- rbinom(n = 1, size = 1, prob = S[6])
      meta_comm[i,6] <- ifelse(meta_comm[i,6] == 1, K[6],0)
    }
  }
  meta_comm$Patch <- c('Patch1','Patch2','Patch3','Patch4','Patch5')
  colnames(meta_comm) <- c('Spp1','Spp2','Spp3','spp4','spp5','Spp6','PatchID')
  beta_list[[n]] <- mean(betadiver(meta_comm[,1:6], method = 'w'))
  nestedness_list[[n]] <- nestedtemp(comm = meta_comm[,1:6])[7]
  meta_comm_list[[n]] <- meta_comm
}
```

---

class: center, middle

# And voila! Just like that we have 100 simulated meta-communities.
# Now, let's get in to the epidiomoligcal modeling!
